Complex Analysis of Askaryan Radiation: UHE-$\nu$ Identification and Reconstruction via the Hilbert Envelope of Observed Signals

### 2025-09-25 template.tex
finished 100 PeV energy recon table

\begin{abstract}
The detection of ultra-high energy neutrinos (UHE-$\nu$), with enegies above 10 PeV, has been a long-time goal in astroparticle physics.  Autonomous, radio-frequency (RF) UHE-$\nu$ detetectors have been deployed in polar regions.  These detectors rely on the Askaryan effect in ice for the neutrino signal.  The Askaryan effect occurs when the excess negative charge within a high-energy cascade radiates in a dense medium.  UHE-$\nu$ can induce such cascades that radiate in the RF bandwidth above thermal backgrounds.  To identify UHE-$\nu$ signals in future data from Askaryan-class detectors, analytic models of the Askaryan electromagnetic field have been created and compared to simulations and laboratory measurements.  These models have correctly described the Askaryan electromagnetic field, while leaving the effect of the RF detection channels on the field to simulation packages.  In this work, we present a fully analytic Askaryan model that accounts for the effect of an RF detection channel.  First, we derive formulas for the observed voltage trace, and for the Hilbert envelope of the trace.  Second, we match the analytic model to data computed for a detector with a single string of eight RF detection channels in NuRadioMC, a key Monte Carlo toolset in UHE-$\nu$ detection.  For UHE-$\nu$ signals, we find correlation coefficients between the analytic signal envelope and MC data in excess of $0.94$, and 99.99\% of UHE-$\nu$ signals pass a correlation threshold of 0.4.  Integrating the normalized PDF of the correlation distribution for RF thermal noise reveals that just 0.2 thermal background events pass the correlation threshold in 5 years of continuous operation at a 1 Hz thermal trigger rate.  Finally, we reconstruct the logarithm of the UHE-$\nu$ cascade energy.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug18_plot1.pdf}
\caption{\label{fig:fig5} (Black circles) Viewing angle from NuRadioMC. (Solid black line) Gaussian fit, with $\mu = 54.8$ deg, and $\sigma = 2.96$ deg.  (Black dashed line) Cherenkov angle, $\theta_{\rm C}$.}
\end{figure}

\item Equations \ref{eq:uncert}-\ref{eq:a_err} may be used to reconstruct the natural logarithm of the UHE-$\nu$ cascade energy, $\ln\Lambda$.  For Eq. \ref{eq:uncert}, $\sigma_t$ is measured from the optimized analytic envelope, $c$ and $\theta_{\rm C}$ are known constants, and an assumption must be made for $\Delta\theta$.  We will make the assumption that $\Delta\theta \approx \Delta\theta_{\rm rms}$.  The assumption is justified since the $\Delta\theta$ distribution is Gaussian.  The MC truth for $\Delta\theta$ in this analysis is shown in Fig. \ref{fig:fig5}, with a Gaussian fit ($\mu=54.8$ deg, $\sigma=2.96$ deg).  The rms and the $\sigma$ parameter are equal for Gaussian distributions with zero mean, so we let $\Delta\theta \to \Delta \theta_{\rm rms}$.  The fractional error $\sigma_{\Delta\theta}/\Delta\theta$ is then set to 1.0, reflecting uncertainty in this parameter.  Solving Eq. \ref{eq:uncert} for $a$ gives

@@

Using Eqs. 10 and 12 from HH2022 (\cite{PhysRevD.105.123019}), $c_{\rm em}$ and $c_{\rm had}$ were found to be 0.80 and 0.93 meters, respectively (FWHM, $R=0.5$).  Using Eq. \ref{eq:lnLambda}, the $\sigma_t$ results from the optimized envelope fits to UHE-$\nu$ CSWs may be used to deduce the logarithm of the UHE-$\nu$ cascade, $\log_{10} E_{\rm C}$.  First, converting to the base-10 logarithm introduces a factor of $\ln(10)$ in the denominator of Eq. \ref{eq:lnLambda}.  Second, $\ln\Lambda = \ln(E_{\rm C}/E_{\rm crit})$, where $E_{\rm C}$ is the cascade energy, and $E_{\rm crit}\approx 10^8$ eV is known as the critical energy \cite{PhysRevD.105.123019}.  Since $\ln\Lambda = \ln E_{\rm C} - \ln E_{\rm crit}$, separating this ratio adds a constant to the right hand side of Eq. \ref{eq:lnLambda}.  The modified form of Eq. \ref{eq:lnLambda} is

\begin{equation}
\log_{10} E_{\rm C} = \frac{\left( c\sigma_t \right)^2}{\ln 10(c_{\rm em/had} \Delta\theta_{\rm rms}\sin\theta_{\rm C})^2} + \log_{10} E_{\rm crit} \label{eq:lnLambda_2}
\end{equation}

Table \ref{tab:2} contains the results for the energy reconstruction.  Note that when applying the change of base formula, factors of $\ln(10)$ cancel in the error ratio of Eq. \ref{eq:a_err_2}, so the error in $\log_{10} E_{\rm C}$ remains $2.0$.

\begin{table}
\centering
\begin{tabular}{| c | c |}
\hline
\textbf{MC Truth $\log_{10} E_{\rm C}$} & \textbf{Reconstructed $\log_{10} E_{\rm C}$} \\
17.0 & $17.8 \pm 2.0$ \\
\hline
\end{tabular}
\caption{\label{tab:2}}
\end{table}

>>>>

\begin{abstract}
The detection of ultra-high energy neutrinos (UHE-$\nu$), with enegies above 10 PeV, has been a long-time goal in astroparticle physics.  Autonomous, radio-frequency (RF) UHE-$\nu$ detetectors have been deployed in polar regions.  These detectors rely on the Askaryan effect in ice for the neutrino signal.  The Askaryan effect occurs when the excess negative charge within a high-energy cascade radiates in a dense medium.  UHE-$\nu$ can induce such cascades that radiate in the RF bandwidth above thermal backgrounds.  To identify UHE-$\nu$ signals in future data from Askaryan-class detectors, analytic models of the Askaryan electromagnetic field have been created and compared to simulations and laboratory measurements.  These models have correctly described the Askaryan electromagnetic field, while leaving the effect of the RF detection channels on the field to simulation packages.  In this work, we present a fully analytic Askaryan model that accounts for the effect of an RF detection channel.  First, we derive formulas for the observed voltage trace, and for the Hilbert envelope of the trace.  Second, we match the analytic model to data computed for a detector with a single string of eight RF detection channels in NuRadioMC, a key Monte Carlo toolset in UHE-$\nu$ detection.  For UHE-$\nu$ signals, we find correlation coefficients between the analytic signal envelope and MC data in excess of $0.94$, and 99.99\% of UHE-$\nu$ signals pass a correlation threshold of 0.4.  Integrating the normalized PDF of the correlation distribution for RF thermal noise reveals that just 0.2 thermal background events pass the correlation threshold in 5 years of continuous operation at a 1 Hz thermal trigger rate.  Finally, we present a preliminary reconstruction of the UHE-$\nu$ cascade energy.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug18_plot1.pdf}
\caption{\label{fig:fig5} (Black circles) Viewing angle from NuRadioMC. (Solid black line) Gaussian fit, with $\mu = -1.05$ deg, and $\sigma = 3.02$ deg.  (Black dashed line) Cherenkov angle, $\theta_{\rm C}$.}
\end{figure}

\item Equations \ref{eq:uncert}-\ref{eq:a_err} may be used to reconstruct the natural logarithm of the UHE-$\nu$ cascade energy, $\ln\Lambda$.  For Eq. \ref{eq:uncert}, $\sigma_t$ is measured from the optimized analytic envelope, $c$ and $\theta_{\rm C}$ are known constants, and an assumption must be made for $\Delta\theta$.  We will make the assumption that $\Delta\theta \approx \Delta\theta_{\rm rms}$.  The assumption is justified since the $\Delta\theta$ distribution is Gaussian.  The MC truth for $\Delta\theta$ in this analysis is shown in Fig. \ref{fig:fig5}, with a Gaussian fit ($\mu=-1.05$ deg, $\sigma=3.02$ deg).  The rms and the $\sigma$ parameter are equal for Gaussian distributions with zero mean, so we let $\Delta\theta \to \Delta \theta_{\rm rms}$.  The fractional error $\sigma_{\Delta\theta}/\Delta\theta$ is then set to 1.0, reflecting uncertainty in this parameter.  Solving Eq. \ref{eq:uncert} for $a$ gives

@@

Using Eqs. 10 and 12 from HH2022 (\cite{PhysRevD.105.123019}), $c_{\rm em}$ and $c_{\rm had}$ were found to be 0.80 and 0.93 meters, respectively (FWHM, $R=0.5$).  Using Eq. \ref{eq:lnLambda}, the $\sigma_t$ results from the optimized envelope fits to UHE-$\nu$ signal CSWs may be used to deduce the logarithm of the UHE-$\nu$ cascade, $\log_{10} E_{\rm C}$.  First, converting to the base-10 logarithm introduces a factor of $\ln(10)$ in the denominator of Eq. \ref{eq:lnLambda}.  Second, $\ln\Lambda = \ln(E_{\rm C}/E_{\rm crit})$, where $E_{\rm C}$ is the cascade energy, and $E_{\rm crit}\approx 10^8$ eV is known as the critical energy \cite{PhysRevD.105.123019}.  Since $\ln\Lambda = \ln E_{\rm C} - \ln E_{\rm crit}$, separating this ratio adds a constant to the right hand side of Eq. \ref{eq:lnLambda}.  Third, let $c_{\rm ave}$ be the average of $c_{\rm em}$ and $c_{\rm had}$, reflecting the unknown hadronic or electromagnetic nature of the cascade.  The modified form of Eq. \ref{eq:lnLambda} is

\begin{equation}
\log_{10} E_{\rm C} = \frac{\left( c\sigma_t \right)^2}{\ln 10(c_{\rm ave} \Delta\theta_{\rm rms}\sin\theta_{\rm C})^2} + \log_{10} E_{\rm crit} \label{eq:lnLambda_2}
\end{equation}

Table \ref{tab:2} contains the preliminary results for the reconstruction of $\log_{10}E_{\rm C}$.  The results are consistent with the MC truth.  Note that when applying the change of base formula, factors of $\ln(10)$ cancel in the error ratio of Eq. \ref{eq:a_err_2}, so the error in $\log_{10} E_{\rm C}$ remains $2.0$.  The $sigma_t$ result is the average value of 15133 optimized analytic envelopes fit to signal CSWs generated by 100 PeV UHE-$\nu$ events that triggered the detector.  Note that the error in the \textit{average} $\sigma_t$ over all events fluctuates due to varying $\Delta\theta$, while the error in \textit{individual} $\sigma_t$ per CSW fit is negligible if the samples per CSW is much greater than 1.  By far, the largest sources of systematic error are \textit{reflected signals} in the CSWs.

Reflected signals are caused when more than one ray-tracing solution is available between the Askaryan radiation source and the detector.  Signals that propagate directly to the detector are known as \textit{direct signals}, while signals that first reflect at the snow-air interface, or in the glacial firn, are known as \textit{reflected signals} \cite{}.  When \textit{direct} and \textit{reflected} signals arrive at the detector at similar times, and one or the other triggers the detector, the recorded waveforms contain direct and reflected signals.

\begin{table}
\centering
\begin{tabular}{| c | c |}
\hline
\textbf{Parameter} & \textbf{Average Value} \\ \hline
MC Truth, $E_{\rm C}$ [eV] & $10^{17}$ \\
MC Truth, $\log_{10}E_{\rm C}$ & $17$ \\
$\sigma_t$ & $1.2 \pm 0.4$ ns \\
$c_{\rm em,had}$ & 0.86 m \\
$\Delta\theta_{\rm rms}$ & 3.9 degrees \\
$E_{\rm crit}$ & 100 MeV \\
$c$ & $0.3/1.78$ m ns$^{-1}$ \\
$\sin\theta_{\rm C}$ & 0.8271 \\ \hline
\textbf{Reconstructed} $\mathbf{\log_{10}E_{\rm C}}$ & $16.5\pm 2.0$ \\
\hline
\end{tabular}
\caption{\label{tab:2} Energy reconstruction parameters used to calculate the logarithm of the UHE-$\nu$ cascade energy, $\log_{10}E_{\rm C}$.}
\end{table}
